---
title: "Exploring Health and Economic Consequences of Severe Weather in U.S."
author: "Ihar Kukharchuk"
date: "April 7, 2018"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

```{r setup, include = FALSE}
library(plyr)
library(R.utils)
library(ggplot2)
library(gridExtra)
# for fast reading
library(data.table)

## just to force appropriate 'locale'
Sys.setlocale(category = "LC_ALL", locale = "C")

## configure plots template: size and location
knitr::opts_template$set(plots = list(fig.align = 'center',
                                      fig.height = 4,
                                      fig.width = 7,
                                      fig.path = 'figures/',
                                      dev = 'png'))
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

```{r, ref.label = "paper-constants", include = FALSE}
```
```{r, ref.label = "obtain-data", include = FALSE}
```
```{r, ref.label = "clean-data", include = FALSE}
```
```{r, ref.label = "health-damage", include = FALSE}
```
```{r, ref.label = "health-damage-chart", include = FALSE}
```
```{r, ref.label = "eco-damage", include = FALSE}
```
```{r, ref.label = "eco-damage-chart", include = FALSE}
```

### Abstract
The exploration of statistical data of severe weather in U.S. helps to identify
which regions require more attention to this problem. In order to reveal this
information the U.S. National Oceanic and Atmospheric Administration's (NOAA)
storm database is using in this analysis. This paper analyses the casualties of
specific events across the U.S.: fatalities, injuries, property and crop damage.
According to this paper tornado is the most harmful event for public health.
Flood and typhoon events cause the biggest negative impact to the economic
of region.

### Exploration of Public Health Damage

Next histogram shows `r max_events_to_display` most harmful events for
public health:

```{r health-damage-chart-repeat, opts.label = 'plots', echo = FALSE}
health_chart
health_dmg
```

The biggest damage is coming from tornado, then excessive heat and tstm wind.

### Exploration of Economic Damage

Next histogram shows `r max_events_to_display` most harmful events for
economic of regions:

```{r eco-damage-chart-repeat, opts.label = 'plots', echo = FALSE}
eco_chart
eco_dmg
```

The biggest economic damage is coming from flood, then hurricane and tornado.

### Algorithm of exploration NOAA storm database for this research

NOAA storm database contains a lot of severe wheather types. There is no
necessity to display every possible event because most of them didn't cause
health or economic consequences. This is the reason why across the research
only `r max_events_to_display` most harmful events are being displayed.

```{r paper-constants, eval = FALSE}
max_events_to_display <- 15
```

Separate function `obtain_data` is added to check the presence of RAW data
locally, if data isn't present it will be loaded from the internet.

```{r obtain-data, eval = FALSE}
## obtain the data
obtain_data <- function(filename, archive_filename, url_location) {
        if (!file.exists(filename) && !file.exists(archive_filename)) {
                download.file(url_location, archive_filename, method = "auto")
        }
        if (!file.exists(filename) && file.exists(archive_filename)) {
                bunzip2(archive_filename, filename, remove = FALSE, skip = TRUE)
        }
        if (!file.exists(filename)) {
                stop("storm data is unavailable: cannot process data, stop")
        }
}

## storm data sources
storm_data_file <- "StormData.csv"
storm_archive_file <- "StormData.csv.bz2"
storm_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

obtain_data(storm_data_file, storm_archive_file, storm_url)

## usual reading
# storm_data <- read.csv(storm_data_file)
## fast reading
storm_data <- fread(storm_data_file)

storm_data_orig <- storm_data
```

The most interesting data for analysing is located in the following variables:

* EVTYPE - type of event
* FATALITIES - the amount of fatalities
* INJURIES - the amount of injuries
* PROPDMG - property damage cost
* PROPDMGEXP - property damage value multiplier
* CROPDMG - crop damage cost
* CROPDMGEXP - crop damage value multiplier

As far as there is no clean representation of damage values it is necessary to
transform it to separate dedicated value to perform analysis:

* PROPDMG and PROPDMGEXP to PROPDMGGEN
* CROPDMG and CROPDMGEXP to CROPDMGGEN

In order to perform this adjustment
[convertion algorithm](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html)
was used. This algorithm suggest juxtaposition between 'EXP' and numeric values.
In order to safely use this algorithm it is important to prove that there is no
change between article author's data and NOAA database which is used on this
paper. To do this the `unique` command was performed under the same columns.
The unique 'EXP' values are the same.

```{r}
unique(storm_data_orig$PROPDMGEXP)
```
```{r}
unique(storm_data_orig$CROPDMGEXP)
```

The following clean and preparation section was used to clean data before
using it in this paper.

```{r clean-data, eval = FALSE}
## clean and prepare the data
## convert property damage values using PROPDMG and PROPDMGEXP values
exp_names <- c("K", "M",  "", "B", "m", "+", "0", "5", "6", "?",
               "4", "2", "3", "h", "7", "H", "-", "1", "8")
exp_values <- c(1e3, 1e6, 1e0, 1e9, 1e6, 1e0, 1e0, 1e5, 1e6, 1e0,
                1e4, 1e2, 1e3, 1e0, 1e7, 1e2, 1e0, 1e1, 1e8)
genuine_prop_dmg <- mapvalues(storm_data$PROPDMGEXP, exp_names, exp_values)
storm_data$PROPDMGGEN <- as.numeric(genuine_prop_dmg) * storm_data$PROPDMG

## convert crop damage values using CROPDMG and CROPDMGEXP values
exp_names <- c( "", "M", "K", "m", "B", "?", "0", "k", "2")
exp_values <- c(1e0, 1e6, 1e3, 1e6, 1e9, 1e0, 1e0, 1e3, 1e2)
genuine_crop_dmg <- mapvalues(storm_data$CROPDMGEXP, exp_names, exp_values)
storm_data$CROPDMGGEN <- as.numeric(genuine_crop_dmg) * storm_data$CROPDMG

storm_data <- storm_data[, c("EVTYPE",
                             "FATALITIES", "INJURIES",
                             "PROPDMGGEN", "CROPDMGGEN")]
```

All fatalities and injuries were summarized in order to provide chart of total
health damage caused by severe wheather events.

```{r health-damage, eval = FALSE}
## 1. Across the United States, which types of events (as indicated in the
##    EVTYPE variable) are most harmful with respect to population health?
storm_data$HEALTHTOTAL <- storm_data$FATALITIES + storm_data$INJURIES

health_dmg <- aggregate(HEALTHTOTAL ~ EVTYPE, storm_data, sum)
health_dmg <- health_dmg[order(health_dmg$HEALTHTOTAL,
                               decreasing = TRUE),][1:max_events_to_display,]
```

```{r health-damage-chart, opts.label = 'plots'}
health_chart <- ggplot(health_dmg, aes(reorder(EVTYPE, -HEALTHTOTAL),
                                       HEALTHTOTAL, label = HEALTHTOTAL)) +
        geom_col() +
        xlab("type of event") +
        ylab("amount of people") +
        ggtitle("The most harmful events for public health (summary)") +
        theme(plot.title = element_text(hjust = 0.5),
              axis.text.x = element_text(angle = 90, hjust = 1))
health_chart
```
```{r}
health_dmg
```

The next chart helps to determine which part of all damage belongs to either
fatalities or injuries health damage

```{r health-damage-allotment}
health_allo_dmg <- aggregate(cbind(FATALITIES, INJURIES, HEALTHTOTAL) ~ EVTYPE,
                             storm_data, sum)
health_allo_dmg <- health_allo_dmg[
        order(health_allo_dmg$HEALTHTOTAL,
              decreasing = TRUE),][1:max_events_to_display,]
health_allo_dmg_data <- health_allo_dmg[, c("FATALITIES", "INJURIES")]
health_allo_dmg_data <- data.frame(rows = health_allo_dmg$EVTYPE,
                                   stack(health_allo_dmg_data))
health_allo_dmg_data
```

```{r health-damage-allotment-chart, opts.label = 'plots'}
health_allo_chart <- ggplot(health_allo_dmg_data,
                            aes(reorder(rows, -values),
                                values,
                                label = rows,
                                group = ind,
                                fill = ind)) +
        geom_col() +
        xlab("type of event") +
        ylab("amount of people") +
        ggtitle("The most harmful events for public health (summary)") +
        theme(plot.title = element_text(hjust = 0.5),
              axis.text.x = element_text(angle = 90, hjust = 1),
              legend.title = element_blank()) +
        scale_fill_discrete(labels = c("Fatalities", "Injuries"))
health_allo_chart
```

The following two charts shows the most harmful events for health - fatalities
and crop separately:

```{r health-damage-separately}
fatalities_dmg <- aggregate(FATALITIES ~ EVTYPE, storm_data, sum)
fatalities_dmg <- fatalities_dmg[
        order(fatalities_dmg$FATALITIES,
              decreasing = TRUE),][1:max_events_to_display,]
fatalities_dmg

injuries_dmg <- aggregate(INJURIES ~ EVTYPE, storm_data, sum)
injuries_dmg <- injuries_dmg[
        order(injuries_dmg$INJURIES,
              decreasing = TRUE),][1:max_events_to_display,]
injuries_dmg
```

```{r health-damage-separately-chart, opts.label = 'plots', fig.height = 8}
fatalities_chart <- ggplot(fatalities_dmg,
                           aes(reorder(EVTYPE, -FATALITIES),
                               FATALITIES, label = FATALITIES)) +
        geom_col() +
        xlab("type of event") +
        ylab("amount of people") +
        ggtitle("The most harmful events for health:\nfatalities") +
        theme(plot.title = element_text(hjust = 0.5),
              axis.text.x = element_text(angle = 90, hjust = 1))
injuries_chart <- ggplot(injuries_dmg,
                         aes(reorder(EVTYPE, -INJURIES),
                             INJURIES, label = INJURIES)) +
        geom_col() +
        xlab("type of event") +
        ylab("amount of people") +
        ggtitle("injuries") +
        theme(plot.title = element_text(hjust = 0.5),
              axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(fatalities_chart, injuries_chart, nrow = 2)
```

All property and crop costs were summarized in order to provide chart of total
economic damage caused by severe wheather events.

```{r eco-damage, eval = FALSE}
## 2. Across the United States, which types of events have the greatest
##    economic consequences?
storm_data$ECOTOTAL <- storm_data$PROPDMGGEN + storm_data$CROPDMGGEN

eco_dmg <- aggregate(ECOTOTAL ~ EVTYPE, storm_data, sum)
eco_dmg <- eco_dmg[order(eco_dmg$ECOTOTAL,
                         decreasing = TRUE),][1:max_events_to_display,]
```

```{r eco-damage-chart, opts.label = 'plots'}
eco_chart <- ggplot(eco_dmg, aes(reorder(EVTYPE, -ECOTOTAL),
                                 ECOTOTAL, label = ECOTOTAL)) +
        geom_col() +
        xlab("type of event") +
        ylab("damage costs") +
        ggtitle("The most harmful events for economic (summary)") +
        theme(plot.title = element_text(hjust = 0.5),
              axis.text.x = element_text(angle = 90, hjust = 1))
eco_chart
```
```{r}
eco_dmg
```

The next chart helps to determine which part of all damage belongs to either
crop or property damage

```{r eco-damage-allotment}
eco_allo_dmg <- aggregate(cbind(PROPDMGGEN, CROPDMGGEN, ECOTOTAL) ~ EVTYPE,
                          storm_data, sum)
eco_allo_dmg <- eco_allo_dmg[
        order(eco_allo_dmg$ECOTOTAL,
              decreasing = TRUE),][1:max_events_to_display,]
eco_allo_dmg_data <- eco_allo_dmg[, c("PROPDMGGEN", "CROPDMGGEN")]
eco_allo_dmg_data <- data.frame(rows = eco_allo_dmg$EVTYPE,
                                stack(eco_allo_dmg_data))
eco_allo_dmg_data
```

```{r eco-damage-allotment-chart, opts.label = 'plots'}
eco_allo_chart <- ggplot(eco_allo_dmg_data,
                         aes(reorder(rows, -values), values,
                             label = rows, group = ind, fill = ind)) +
        geom_col() +
        xlab("type of event") +
        ylab("damage costs") +
        ggtitle("The most harmful events for economic (summary)") +
        theme(plot.title = element_text(hjust = 0.5),
              axis.text.x = element_text(angle = 90, hjust = 1),
              legend.title = element_blank()) +
        scale_fill_discrete(labels = c("Property", "Crop"))
eco_allo_chart
```

The following two charts shows the most harmful events for economic for
property and crop separately:

```{r eco-damage-separately}
prop_dmg <- aggregate(PROPDMGGEN ~ EVTYPE, storm_data, sum)
prop_dmg <- prop_dmg[order(prop_dmg$PROPDMGGEN,
                           decreasing = TRUE),][1:max_events_to_display,]
prop_dmg

crop_dmg <- aggregate(CROPDMGGEN ~ EVTYPE, storm_data, sum)
crop_dmg <- crop_dmg[order(crop_dmg$CROPDMGGEN,
                           decreasing = TRUE),][1:max_events_to_display,]
crop_dmg
```

```{r eco-damage-separately-chart, opts.label = 'plots', fig.height = 8}
prop_chart <- ggplot(prop_dmg, aes(reorder(EVTYPE, -PROPDMGGEN),
                                   PROPDMGGEN, label = PROPDMGGEN)) +
        geom_col() +
        xlab("type of event") +
        ylab("damage costs") +
        ggtitle("The most harmful events for economic:\nproperty damage") +
        theme(plot.title = element_text(hjust = 0.5),
              axis.text.x = element_text(angle = 90, hjust = 1))
crop_chart <- ggplot(crop_dmg, aes(reorder(EVTYPE, -CROPDMGGEN),
                                   CROPDMGGEN, label = CROPDMGGEN)) +
        geom_col() +
        xlab("type of event") +
        ylab("damage costs") +
        ggtitle("crop damage") +
        theme(plot.title = element_text(hjust = 0.5),
              axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(prop_chart, crop_chart, nrow = 2)
```


### Conclusion
Based on NOAA storm database this paper proves that the most harmful event
according to public health is tornado. The most harmful event according to
economic consequences is flood.